Sequence Specific Modeling of \emph{E.~coli} Cell-Free Protein Synthesis

### 2017-03-31 Paper_v1.tex
updated intro

\section{Introduction}
Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized protein production.
Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
This allows us to interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamental biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing, we must first understand the performance limits of these systems.
One tool to address this question is mathematical modeling.

@@

Stoichiometric reconstructions of microbial metabolism, popularized by flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B. subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
In this study, we used sequence specific constraints based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
A core \emph{E.~coli} cell-free metabolic model was developed from literature \cite{Feist:2007aa}.
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of promoter function, transcription and translation processes.
Thus, sequence specific constraints based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight different proteins.
From this, higher carbon number proteins typically had lower productivity rates, energy efficiency, and carbon yields than that of the lower carbon number proteins.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Analysis.pdf}
\caption{CFPS productivity of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Productivity across the ensemble (error bars represent 95\% CI). B. Mean productivity versus carbon number. Single trendline (dotted line) calculated across all cases (y = 6015004x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99). Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Prod}
\end{figure}

\subsubsection{Productivity}
The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prod}).
All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prod}A).
%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.
This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
However, each individual protein had a different level of productivity.
For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prod}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
The relative high productivity of CAT was due to its T7 promoter.
CAT on a P70a promoter frollowed the same trendline as the other proteins and had a productivity of about 8.8 $\mu$M/h.
Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a P70a promoter in CFPS, given the size of the protein.
%The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
% further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.

\subsubsection{Energy efficiency}
The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 32\% when amino acids were not available (Fig.~\ref{fig:Energy}).
Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein.
When amino acids were supplied in the media (first two cases), there was a comparable performance in having the highest energy efficiency, despite the network's ability to synthesis amino acids.
For the case when amino acids were removed from the media, protein production resulted in a low energy efficiency; below 32\% depending on the protein.
This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis and meet the necessary energy demands of CFPS.
We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Energy}B).
The same inverse trend was observed as for productivity, except that it was linear.
%The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
Proteins with a high carbon number had a lower energy efficiency since they have a higher transcription and translation cost then smaller proteins.
The cases supplemented with amino acids had the same trendline whereas when amino acids were not available there was a significant drop in energy efficiency.
Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency of about 28\% regardless of carbon number.
In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.6 $\pm$ 1.1\% compared to the theoretical maximum of 77.3 $\pm$ 7.3\%.
This shows CFPS systems have a lot of room for improvement: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potential could be diverted towards a protein of interest.
Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which is a result of inefficient energy utilization.

@@

The mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Yield}).

@@

A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Yield}B).

@@

Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})
For example, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
A knockout of the Entner-Doudoroff pathway has no effect on the norm productivity of CAT compared to no knockouts (Fig.~\ref{fig:norm}A).
In addition, pairwise knockouts of Entner-Doudoroff and most subgroups in the network result in the same optimal solution of CAT productivity.
However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B), ssFBA will reroute the flux to optimize the objective function.
Interestingly, a single group knockout of glycolysis/gluconoeogenesis, glutamate/glutamine biosynthesis, alanine/aspartate/asparagine biosynthesis was detrimental to CAT productivity.
Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
In our case, ssFBA reached the same optimal solution of CAT productivity for 73\% of the pairwise group knockouts.
To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
\begin{figure}[t!]
\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribition-Analysis.pdf}
\caption{The norm effect of pairwise knockouts of subgroups in the cell-free network (A) CAT productivity norm compared to no knockouts in the experimentally constrained case. (B) Flux distribution norm compared to no knockouts in the experimentally constrained case.}
\label{fig:norm}
\end{figure}

>>>>

\section{Introduction}
Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized protein production.
Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
This allows us to interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamental biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
One tool to address this question is mathematical modeling.

@@

Stoichiometric reconstructions of microbial metabolism, popularized by flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
Constraints based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions. 
These constraints based approaches have predicted productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} of arbitrary complexity, including genome scale networks, using linear programming\cite{Covert:2004aa}.  
Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B. subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
Stoichiometric reconstructions have been expanded to include metabolism integrated with detailed gene expression (ME-Model) \cite{Allen:2003aa, Brien693}, and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220} which has increased the scope of simulating genetic and environmental perturbations.
%However, FBA has been shown to have alternative solutions that lead to the same desired optimal phenotype using flux variability analysis \cite{Mahadevan2003264} or mixed-integer linear programming based algorithms \cite{LEE2000711}. 
Constraints based methods are great tools to estimate the performance of metabolic networks with none to very few parameters. 
However, constraints based methods are typically used to model \textit{in~vivo} processes and have not been applied to cell-free metabolism.

In this study, we used sequence specific constraints based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
A core \emph{E.~coli} cell-free metabolic model was developed from literature \cite{Feist:2007aa}.
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of promoter function, transcription and translation processes.
Thus, sequence specific constraints based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight different proteins.
From this, higher carbon number proteins typically had lower productivity rates, energy efficiency, and carbon yields than that of the lower carbon number proteins.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Analysis.pdf}
\caption{CFPS performance of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Productivity. B. Mean productivity versus carbon number. Single trendline (dotted line) calculated across all cases (y = 6.02$\cdot$10\textsuperscript{6}x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99). 
C. Energy efficiency. D. Mean energy efficiency versus carbon number. Trendline for cases with amino acids (black dotted line; y = -1.65$\cdot$10\textsuperscript{-2}x+90.95; R\textsuperscript{2} = 0.91) and trendline for without amino acids (grey dotted line; y = -3.18$\cdot$10\textsuperscript{-3}x+33.48; R\textsuperscript{2} = 0.78).
E. Carbon yield. F. Mean carbon yield versus carbon number. Trendline for cases with amino acids (black dotted line; y = -8.03$\cdot$10\textsuperscript{-3}x+63.83; R\textsuperscript{2} = 0.95) and trendline for without amino acids (grey dotted line; y = -6.98$\cdot$10\textsuperscript{-3}x+58.86; R\textsuperscript{2} = 0.90).
Error bars: 95\% CI; asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
\label{fig:Prof}
\end{figure}

\subsubsection{Productivity}
The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prof}A).
%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.
This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
However, each individual protein had a different level of productivity.
For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
The relative high productivity of CAT was due to its T7 promoter.
CAT on a P70a promoter frollowed the same trendline as the other proteins and had a productivity of about 8.8 $\mu$M/h.
Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a P70a promoter in CFPS, given the size of the protein.
%The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
% further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.

\subsubsection{Energy efficiency}
The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 32\% when amino acids were not available (Fig.~\ref{fig:Prof}C-D).
Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein.
When amino acids were supplied in the media (first two cases), there was a comparable performance in having the highest energy efficiency, despite the network's ability to synthesis amino acids.
For the case when amino acids were removed from the media, protein production resulted in a low energy efficiency; below 32\% depending on the protein.
This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis and meet the necessary energy demands of CFPS.
We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Prof}D).
The same inverse trend was observed as for productivity, except that it was linear.
%The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
Proteins with a high carbon number had a lower energy efficiency since they have a higher transcription and translation cost then smaller proteins.
The cases supplemented with amino acids had the same trendline whereas when amino acids were not available there was a significant drop in energy efficiency.
Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency of about 28\% regardless of carbon number.
In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.6 $\pm$ 1.1\% compared to the theoretical maximum of 77.3 $\pm$ 7.3\%.
This shows CFPS systems have a lot of room for improvement: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potential could be diverted towards a protein of interest.
Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which is a result of inefficient energy utilization.

@@

The mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).

@@

A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Yield}F).

@@

Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})
For example, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
A knockout of the Entner-Doudoroff pathway has no effect on the norm productivity of CAT compared to no knockouts (Fig.~\ref{fig:norm}A).
In addition, pairwise knockouts of Entner-Doudoroff and other subgroups (i.e. pentose phosphate pathway, cofactors, folate metabolism, etc.) in the network result in the same optimal solution of CAT productivity.
However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B); ssFBA will reroute the flux to optimize the objective function.
Interestingly, there are crucial subgroups that are detrimental to CAT productivity. 
For example, a single group knockout of glycolysis/gluconoeogenesis or oxidative phosphorylation leads to no CAT production. 
There were also cases of group knockouts that had no effect on productivity or its flux distribution such as isoleucine, leucine, and valine biosynthesis knocked out with histidine biosynthesis. 
Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
In our case, ssFBA reached the same optimal solution of CAT productivity for 39.9\% of the pairwise group knockouts, but had a different flux distribution than compared to no knockouts.
To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
\begin{figure}[t!]
\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribition-Analysis.pdf}
\caption{The norm effect of pairwise knockouts of subgroups in the cell-free network (A) CAT productivity norm compared to no knockouts in the experimentally constrained case. (B) Flux distribution norm compared to no knockouts in the experimentally constrained case.}
\label{fig:norm}
\end{figure}
